Gravitational waves from colliding black holes

### 2017-04-09 Report/Draft/draft.tex
Updated report

Now that the existence of gravitational waves has been confirmed and their detection is possible we enter a new era of astronomy, and it is difficult to overstate the wealth of science that is now attainable in the coming decades. Gravitational waves can be used to study astrophysical phenomena that cannot be observed using electromagnetic radiation, such as binary black hole (BBH) systems where two inspiralling black holes merge into one, as well as being used in conjunction with electromagnetic observations in 'multi-messenger' astronomy where events such as supernovae and gamma ray bursts are thought to release both gravitational and electromagnetic radiation which can be studied in conjunction with one another\cite{mm}\cite{mm2}. In addition GWs can be used to conduct the closest tests of general relativity (GR) to date, (CITE CAL's) as well as to further inform the ongoing development of quantum gravity models\cite{qgrav}. Currently the LIGO network consists of two detectors, one in Hanford WA, one in Livingston LA, however there are three detectors that will be added to the network in the coming years, with VIRGO in Italy due to come online by the end of 2017, and with LIGO India and KAGRA joining later. In addition to this, there are also proposals underway for a space-based gravitational wave observatory, the Laster Interferometric Space Antenna (LISA)\cite{lisa} which would be able to explore a different frequency-range and therefore study a range of different astrophysical objects to earth-based detectors.

@@

\subsection{Gravitational waves and their sources}
A complete analysis is available in Hartle\cite{hartle}, but here we briefly overview the fundamental theory of gravitational waves. In the general theory of relativity, gravity is a consequence of the curvature of a 4-dimensional spacetime as described by the Einstein equation:
\begin{equation}
R_{\alpha\beta}-\frac{1}{2}g_{\alpha\beta}R=8\pi T_{\alpha\beta}
\end{equation}
where $R_{\alpha\beta}$ is the Riemannian curvature tensor, $g_{\alpha\beta}$ is the metric tensor, $R$ is the Ricci scalar and $T_{\alpha\beta}$ is the metric tensor. This equation is essentially ten non-linear partial differential equations, where we use the Einstein summation convention. Intuitively, the LHS of this equation can be thought of as the local curvature of spacetime, and the RHS quantifies the energy and momentum density. In the weak-field regime, where the curvature of spacetime is low, the metric tensor can be approximated as
\begin{equation}
g_{\alpha\beta}(x)=\eta_{\alpha\beta}+h_{\alpha\beta}(x).
\end{equation}
where $\eta_{\alpha\beta}$ is the Minkowski metric and $|h_{\alpha\beta}| \ll 1$ for all components. This metric can be substituted into the Einstein equation, and expanding in $h_{\alpha\beta}$ in first order and using the Lorentz gauage, the Einstein equation becomes
\begin{equation}
\square h_{\alpha\beta}=0
\end{equation}
where $\square$ is the D'Alembertian operator, with
\begin{equation}
\partial_\beta h^\beta_\alpha-\frac{1}{2}\partial_\alpha h^\beta_\beta=0.
\end{equation}
The most general solution to this equation is

\begin{equation}
h_{\alpha\beta}=
\begin{pmatrix}
0 & 0 & 0 & 0\\
0 & a_+ & a_x & 0\\
0 & a_x & -a_+ & 0\\
0 & 0 & 0 & 0\\
\end{pmatrix}
e^{i\omega(z-t)}
\end{equation}
for a wave with frequency $\omega$ propagating in the $z$ direction. Here the $a_+$ and $a_x$ terms represent the amplitudes of the 'plus' and 'cross' polarisations respectively. As a gravitational wave passes through an observer, spacetime is distorted along the spatial directions orthogonal to the propagation direction of the wave according to these polarisation amplitudes. A visualisation of this is shown in Fig. 1. It is this stretching and squeezing of spacetime that the LIGO detectors were built to detect.

@@

\begin{figure}[h]
	\includegraphics[scale=0.60]{fig2.jpg}

	\centering
	\caption{\cite{fig2}Impression of two inspiralling compact objects emitting gravitational waves}
	\centering
\end{figure}

Approximating that the field around the source is still weak, that the wavelength is long and that the observer is a large distance from the source, the spatial elements of the GW metric are
\begin{equation}
h_{ij}\approx\frac{2}{r}\ddot{I}^{ij}(t-r)
\end{equation}
where $I^{ij}$ is the second mass moment given by
\begin{equation}
I^{ij}(t)=\int d^3x\mu(t,\vec{x})x^jx^j
\end{equation}
where $\mu(t,\vec{x})$ is the mass density of the system. The energy loss of a binary system emitting gravitational waves is
\begin{equation}
L_{GW}=\frac{128}{5}M^2R^4\Omega^6,
\end{equation}
and as the binary loses energy, the separation between the compact objects decreases, increasing the orbital frequency. Given the connection between the mass distribution of the system in (7) and the GW amplitudes in (6), this gives rise to a 'chirp' effect seen in the signal of a CBC GW. This is shown in a Fig. 3. which is a simulated waveform of the merging of a BBH system.
\subsection{LIGO}
Overview of detectors
\subsection{Parameter estimation and Bayesian inference}
Bayes theorem and how it is used in PE
\subsection{MCMC}
Overview of MCMCs that will be used for inference
\subsection{Precession}
Precessing binaries and effective spin parameters
\subsection{Astrophysical implications}
Binary formation models and how estimating chi p is significant for distinguishing bewteen them

>>>>

Now that the existence of gravitational waves has been confirmed and their detection is possible we enter a new era of astronomy, and it is difficult to overstate the wealth of science that is now attainable in the coming decades. Gravitational waves can be used to study astrophysical phenomena that cannot be observed using electromagnetic radiation, such as binary black hole (BBH) systems where two inspiralling black holes merge into one, as well as being used in conjunction with electromagnetic observations in 'multi-messenger' astronomy where events such as supernovae and gamma ray bursts are thought to release both gravitational and electromagnetic radiation which can be studied in conjunction with one another\cite{mm}\cite{mm2}. In addition GWs can be used to conduct the closest tests of general relativity (GR) to date, (CITE CAL's) as well as to further inform the ongoing development of quantum gravity models\cite{qgrav}. Currently the LIGO network consists of two detectors, one in Hanford WA (H1), one in Livingston LA (L1), however there are three detectors that will be added to the network in the coming years, with VIRGO (V1) in Italy due to come online by the end of 2017, and with LIGO India and KAGRA joining later. In addition to this, there are also proposals underway for a space-based gravitational wave observatory, the Laster Interferometric Space Antenna (LISA)\cite{lisa} which would be able to explore a different frequency-range and therefore study a range of different astrophysical objects to earth-based detectors.

@@

\subsection{Gravitational waves and their sources}
A complete analysis is available in Hartle\cite{hartle}, but here we briefly overview the fundamental theory of gravitational waves. In the general theory of relativity, gravity is a consequence of the curvature of a 4-dimensional spacetime as described by the Einstein equation (in natural units):
\begin{equation}
R_{\alpha\beta}-\frac{1}{2}g_{\alpha\beta}R=8\pi T_{\alpha\beta}
\end{equation}
where $R_{\alpha\beta}$ is the Riemannian curvature tensor, $g_{\alpha\beta}$ is the metric tensor, $R$ is the Ricci scalar and $T_{\alpha\beta}$ is the energy-momentum tensor. This equation is essentially ten non-linear partial differential equations, where we use the Einstein summation convention to sum over all indices, and where indices run from $0$ to $4$. Intuitively, the LHS of this equation can be thought of as the local curvature of spacetime, and the RHS quantifies the energy and momentum density. In the weak-field regime, where the curvature of spacetime is low, the metric tensor can be approximated as
\begin{equation}
g_{\alpha\beta}(x)=\eta_{\alpha\beta}+h_{\alpha\beta}(x).
\end{equation}
where $\eta_{\alpha\beta}$ is the Minkowski metric and $|h_{\alpha\beta}| \ll 1$ for all components. This metric can be substituted into the Einstein equation, and expanding in $h_{\alpha\beta}$ in first order and using the Lorentz gauage, the Einstein equation becomes
\begin{equation}
\square h_{\alpha\beta}=0
\end{equation}
where $\square$ is the D'Alembertian operator, with
\begin{equation}
\partial_\beta h^\beta_\alpha-\frac{1}{2}\partial_\alpha h^\beta_\beta=0.
\end{equation}
The general solution to this equation is

\begin{equation}
h_{\alpha\beta}=
\begin{pmatrix}
0 & 0 & 0 & 0\\
0 & a_+ & a_\times & 0\\
0 & a_\times & -a_+ & 0\\
0 & 0 & 0 & 0\\
\end{pmatrix}
e^{i\omega(z-t)}
\end{equation}
for a wave with frequency $\omega$ propagating in the $z$ direction. Here the $a_+$ and $a_\times$ terms represent the amplitudes of the 'plus' and 'cross' polarisations respectively. As a gravitational wave passes through an observer, spacetime is distorted along the spatial directions orthogonal to the propagation direction of the wave according to these polarisation amplitudes. A visualisation of this is shown in Fig. 1. It is this stretching and squeezing of spacetime that the LIGO detectors were built to detect.

@@

\begin{figure}[h]
	\includegraphics[scale=0.60]{fig2.jpg}
	\centering
	\caption{\cite{fig2}Impression of two inspiralling compact objects emitting gravitational waves}
	\centering
\end{figure}

Approximating that the field around the source is still weak, that the wavelength is long and that the observer is a large distance from the source, the spatial elements of the GW metric are
\begin{equation}
h_{ij}\approx\frac{2}{r}\ddot{I}^{ij}(t-r)
\end{equation}
where $I^{ij}$ is the second mass moment given by
\begin{equation}
I^{ij}(t)=\int d^3x\mu(t,\vec{x})x^jx^j
\end{equation}
where $\mu(t,\vec{x})$ is the mass density of the system. The energy loss of a binary system emitting gravitational waves is
\begin{equation}
L_{GW}=\frac{128}{5}M^2R^4\Omega^6,
\end{equation}
and as the binary loses energy, the separation between the compact objects decreases, increasing the orbital frequency. Given the connection between the mass distribution of the system in (7) and the GW amplitudes in (6), this gives rise to a 'chirp' effect seen in the signal of a CBC GW. This is shown in a Fig. 3. which is a simulated waveform of the merging of a BBH system.

\begin{figure}[h]
	\includegraphics[scale=0.60]{fig3.png}
	\centering
	\caption{Simulated waveform of a BBH merger of two 35 solar mass black holes.}
	\centering
\end{figure}

The waveform can be split into three phases - the inspiral, the merger and the ringdown. The frequency, frequency evolution and amplitudes of each polarisation of the GW emitted from a merger will depend on the properties of the system itself, and as such there is information about the source contained in the specific morphology of a GW signal.

\subsection{LIGO}
\begin{figure}
	\includegraphics[scale=0.35]{fig4.png}
	\centering
	\caption{Simplified schematic of a LIGO detector http://images.iop.org/objects/phw/news/20/2/10/ligo-figure.png}
	\centering
	\includegraphics[scale=0.60]{fig5.png}
	\centering
	\caption{\cite{noise3}Noise spectrum for advanced LIGO detectors.}
	\centering
\end{figure}
The fundamental physical principle of the LIGO detectors is that they are laser interferometeres. Interferometers are devices that can measure changes in length to extremely high accuracy using constructive and destructive interference, as shown in Fig. 4. Light leaves the laser beam, and is split down the two arms of the detector by the beam splitter. If the length of the two arms is equal, the optical path difference between the two light beams is zero, and the two beams constructively interfere after recombining at the beam splitter. However if there is any change in the length of one of the arms, the beams will no longer constructively interfere and the photodetector will register a change in intensity. As a gravitational wave passes through the detector, the relative length of the arms changes, and the GW signal is recorded by the photodiode. A large part of the scientific and engineering effort at LIGO involves techniques to minimise and account for noise in the system, and the recent advanced LIGO upgrade to the detectors increased the effective volume within which mergers can be detected by a factor of 3\cite{noise}\cite{noise2}. Some of these techniques include suspending the mirrors from a series of pulleys and penduli, and applying real time corrections to the positions of the mirrors to compensate for external seismic noise. There is also considerable effort in managing the optics and lasers of the system in order to maximise the coherence of the laser beam and the intensity detected in the photodiode\cite{lasers}. The noise spectrum for advanced LIGO is shown in Fig. 5.\cite{noise3}, where the sharp noise peaks are specifically designed resonances that are extracted out of the strain data. The strain data observed in the detector is a function of the different polarisation amplitudes
\begin{equation}
h(t)=F^{+}(\alpha,\delta,\psi)h_+(t)+F^{\times}(\alpha,\delta,\psi)h_\times(t),
\end{equation}
where $F^{+}(\alpha,\delta,\psi)$ and $F^{x}(\alpha,\delta,\psi)$ are the antenna beam patterns that describe how the detector responds to signals at different sky locations and polarisations\cite{beampat}. In first order, the polarisations are given by
\begin{equation}
h_+(t)=A_{GW}(t)(1+cos^2(\iota)cos(\phi(t)))
\end{equation}
\begin{equation}
h_\times(t)=-2A_{GW}(t)cos(\iota)sin(\phi(t))
\end{equation}
and binaries that are face on (with $\iota=0$) emit circularly polarised waves, and edge-on binaries emit linearly (either cross or plus) polarised GWs.

\subsection{Parameter estimation and Bayesian inference}
A signal is described by a total of 16 parameters\cite{props} - time and phase of coalesence $t_c$ and $\phi_c$, two parameters to describe sky location (right ascension, $\alpha$ and declination $\delta$), distance $d$, inclination angle $\iota$ describing the orientation of the binary's total angular momentum with respect to the line of sight, polarisation angle $\psi$, the masses $m_1$, $m_2$, six spin parameters to totally describe the spins on each of the two black holes $\vec{S}_1$, $\vec{S}_2$, and then two eccentricity parameters.

\subsection{MCMC}
Overview of MCMCs that will be used for inference
\subsection{Precession}
Precessing binaries and effective spin parameters
\subsection{Astrophysical implications}
Binary formation models and how estimating chi p is significant for distinguishing bewteen them
